---
title: "STL analysis of Waterville monitoring data -- Part I: Data Processing"
author: "Song Qian"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

## Introduction

STL analysis of long-term nutrient monitoring data from Heidelberg University. Data retrieved from National Center for Water Quality Research: <https://ncwqr.org/>.

### Front matters

Setting up proper file paths, defining some useful functions, and loading necessary packages

```{r}
#| label: Front_Matter 
#| include: false
#| eval: true
#| echo: false
#| message: false
packages<-function(x, repos="http://cran.r-project.org", ...){
      x<-as.character(match.call()[[2]])
      if (!require(x,character.only=TRUE)){
        install.packages(pkgs=x, repos=repos, ...)
        require(x,character.only=TRUE)
      }
    }

base <- getwd()
dataDIR <- paste(base, "Data", sep="/")
    ## put your data set in the Data subdirectory
plotDIR <- paste(base, "Figs", sep="/")
    ## put created figures in the Figures subdirectory
    ##setwd(base)

packages(lattice)
packages(arm)
packages(zoo)
    ##packages(maps)
packages(tikzDevice)
packages(tidyverse)

packages(foreign)
data.restore (paste(dataDIR, "visualizing.data", sep="/"))

median.polish.ts <- function(data.ts, ylab="", plt=T){
			  # median polishing for missing value imputation
  medpolish(matrix(data.ts, ncol=12, byrow=T), eps=0.001, na.rm=T)->temp.2w
  print(names(temp.2w))
  year.temp <- rep(seq(start(data.ts)[1], end(data.ts)[1]), each=12)
  month.temp <- rep(1:12, length(seq(start(data.ts)[1], end(data.ts)[1])))
					  # plotting median polishing results
  if (plt){
    par(mfrow=c(2,1))
    plot(seq(start(data.ts)[1], end(data.ts)[1]),
    temp.2w$overall+temp.2w$row, type="l",
    xlab="Year", ylab=ylab, main="De-seasonalized Trend")
    plot(seq(1,12), temp.2w$overall+temp.2w$col, type="l",
    xlab="Month", ylab=ylab, main="Seasonal Changes")
  }
  data.ts[is.na(data.ts)]<-temp.2w$overall +
    temp.2w$row[year.temp[is.na(data.ts)]-start(data.ts)[1]+1]+
    temp.2w$col[month.temp[is.na(data.ts)]]
    data.ts[is.na(data.ts)]<-temp.2w$overall
  invisible(data.ts)
}

## water year conversion function
water_year <- function(d) {
  start_year <- ifelse(as.integer(format(d, "%m")) < 10,
    as.integer(format(d, "%Y")) - 1,
    as.integer(format(d, "%Y"))
    )
  water_year <- ifelse(as.integer(format(d, "%m"))>=10,
    as.integer(format(d, "%Y")) + 1,
    as.integer(format(d, "%Y"))
    )
  return (data.frame(water_year=water_year,
                     wy_date=as.integer(
                       difftime(d, as.Date(paste0(start_year,
                                                  "-10-01")),
                                units="day")+1)))
}
```

### Reading Maumee River Data

Including initial data processing

```{R}
#| label: data_import
#| message: false

##wvldata <- as_tibble(read.csv(paste(dataDIR, "maumeedata2023.csv", sep="/"), 
##				 header=T, stringsAsFactors=F, na.strings = "#N/A"))

wvldata <- as_tibble(read.csv(paste(dataDIR, "maumeedata2024.csv", sep="/"), 
		 header=T, stringsAsFactors=F, na.strings = "#N/A"))

## names(wvldata)
names(wvldata)<-c("Date", "QFlow", "FLOW","QTSS","TSS",
         		      "QTP", "TP", "QSRP","SRP","QNO23",
    		          "NO23","QTKN","TKN","QCL","CL","QSO4",
    		          "SO4","QSI","SI","QCOND","COND")  

wvldata[wvldata<0] <-  NA
sum(wvldata$SRP > wvldata$TP, na.rm=T)

wvldata$SRP[wvldata$SRP > wvldata$TP] <- NA
wvldata$SRP[wvldata$SRP > 8] <- NA 
    ## detection limits (Check with Nate)
wvldata$SRP[wvldata$QSRP=="below detection" | 
              wvldata$QSRP=="Below Detection"]<-0.0034
wvldata$SRP[wvldata$SRP<0.0034] <- 0.0034
wvldata$NO23[wvldata$QNO23=="below detection"] <- 0.0057  
wvldata$NO23[wvldata$NO23<0.0057] <- 0.0057
wvldata$Rdate <- as.Date(wvldata$Date, format="%m/%d/%Y %H:%M")
wvldata <- wvldata[1:(dim(wvldata)[1]-1), ]

## importing 2024-2025 data
wvldata24_25 <- as_tibble(read.csv(paste(dataDIR, "maumeedata2025.csv",
                                         sep="/"), 
                                       header=T, stringsAsFactors=F,
                                       na.strings = "#N/A"))

wvldata24_25$Rdate <- as.Date(wvldata24_25$Date, format="%m/%d/%y")
wvldata_addition<- wvldata24_25[wvldata24_25$Rdate > as.Date("2024-10-01"), 
                                c(5,8:21 )]
names(wvldata_addition)
names(wvldata_addition) <- c("Date", "F", "CL", "NO2", "NO3","SO4",
                             "SIO2", "NH3",  "SRP",  "TP",  "TKN",
                             "SS",   "NPOC", "TN",   "COND")  
```

### Processing Dates

```{r}
#| label: dates
  wvldata <- mutate(wvldata, 
  		    Rdate = as.Date(Date, format="%m/%d/%Y %H:%M"),
  		    Month = format(Rdate, format="%m"),
  		    Year = format(Rdate, format="%Y"), 
  		    julian = format(Rdate, format="%j%"),
  		    week = format(Rdate, format="%U"))

```

### Finding missing days

```{r}
#| label: missing days
#| message: false
packages(reshape2)
wvl.molten <- melt(as.data.frame(wvldata),
          		     id=c("Rdate", "Month","julian","Year", "week"),
  		             measure.vars=c("FLOW","TSS","TP","SRP","NO23","TKN")) 

tmp <- wvl.molten$Year > "1981"

wvl.daily <- dcast(wvl.molten[tmp,], Rdate ~ variable, mean)
wvl.daily <- filter(wvl.daily, !is.na(Rdate))
tmp <- range(wvl.daily$Rdate)
date <- seq(tmp[1], tmp[2], 1)
temp.dates <- data.frame(Rdate=date, 
                         Month=ordered(format(date, "%b"), levels=month.abb),
                         julian = format(date, "%j"),
                         Year = format(date, "%Y"),
                         week = format(date, "%U")
                         )

wvl.daily <- merge(x=wvl.daily, y=temp.dates, by="Rdate", all=T)
wvl.daily <- mutate(wvl.daily,
                    yrmn = format(Rdate, format="%Y-%b"),
                    yrwk = format(Rdate, format="%Y-%U"))
##  as_tibble(wvl.daily)
```

## Preparing for missing value imputation

```{R}
#| label: missing_impute_1
    ## When there are more than a (calendar) month of missing values,
    ## we use median polishing
wvl.molten2 <- melt(wvl.daily, id=c("Rdate", "Month","julian", 
                                    "week","Year"), 
                    measure.vars=c("FLOW","TSS","TP","SRP","NO23","TKN"))

wvl.weekly <- dcast(wvl.molten2, Year+Month ~ variable, median, na.rm=T)
year.weeks <- tapply(wvl.weekly$TP, wvl.weekly$Year, length)

##  to2 <- function(x){
##    ## converts a single digit integer `x` to character "0x"
##    return(ifelse (x<10, paste("0",x, sep=""), as.character(x)))
##  }

NAimpute <- function(colN, daily=as.data.frame(wvl.daily), 
                     weekly=wvl.weekly){
  yr.wks <- tapply(weekly[,colN], weekly$Year, length)
  wkly <- matrix(NA, nrow=length(yr.wks), ncol=max(yr.wks))
  for (i in 1:length(yr.wks)){
    for (j in 1:max(yr.wks)){
	temp <- weekly$Year==names(yr.wks)[i] & 
	  as.numeric(weekly$Month)==j
	if (sum(temp)>0)
	  wkly[i,j] <- weekly[temp, colN]
    }
  }
  med <- medpolish(wkly, na.rm=T)
  tmp <- is.na(daily[,colN])
  print(paste("Number of NAs to be imputed:", sum(tmp)))
  if (sum(tmp)>0){
    row.yr <- as.numeric(ordered(daily$Year))
    col.wk <- as.numeric(daily$Month) 
    daily[tmp, colN] <- med$overall + med$row[row.yr[tmp]] +
                        med$col[col.wk[tmp]]    
  }
  return(daily[,colN])
}
```

## Missing value imputation

```{r}
#| label: missing_impute_2
wvl.daily <- group_by(wvl.daily, yrwk) %>%
  mutate(TP = ifelse(is.na(TP), median(TP, na.rm=T), TP),
         SRP = ifelse(is.na(SRP), median(SRP, na.rm=T), SRP),
         FLOW = ifelse(is.na(FLOW), median(FLOW, na.rm=T), FLOW),
         NO23 = ifelse(is.na(NO23), median(NO23, na.rm=T), NO23),
         TKN = ifelse(is.na(TKN), median(TKN, na.rm=T), TKN),
         TSS = ifelse(is.na(TSS), median(TSS, na.rm=T), TSS)) %>%
  group_by(yrmn) %>%
  mutate(TP = ifelse(is.na(TP), median(TP, na.rm=T), TP),
         SRP = ifelse(is.na(SRP), median(SRP, na.rm=T), SRP),
         FLOW = ifelse(is.na(FLOW), median(FLOW, na.rm=T), FLOW),
         NO23 = ifelse(is.na(NO23), median(NO23, na.rm=T), NO23),
         TKN = ifelse(is.na(TKN), median(TKN, na.rm=T), TKN),
         TSS = ifelse(is.na(TSS), median(TSS, na.rm=T), TSS))

  ##  wvl.daily$TP <- NAimpute(colN="TP")
  ##  wvl.daily$SRP <- ifelse(wvl.daily$SRP<0, NA, wvl.daily$SRP)
  ##  wvl.daily$SRP <- NAimpute(col="SRP")
  wvl.daily$FLOW <- NAimpute(colN="FLOW")
  ##  wvl.daily$NO23 <- NAimpute(col="NO23")
  wvl.daily$TKN <- NAimpute(col="TKN")
  ##  wvl.daily$TSS <- NAimpute(col="TSS")
  ##    names(wvl.daily)
  ##    wvl.daily$Yr_Mn <- format(wvl.daily$Rdate, "%m/%Y")
  wvl.daily$TSS <- wvl.daily$TSS + 0.1 ## for 6 0s
```

## Flow Data from USGS

There were at least two consequtive months of flow data missing from Heidelberg. We go directly to USGS to get daily flow data from the Waterville gauge: 04193500

```{r}
#| label: USGS_Flow
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "Comparison of flow data sourced from USGS and Heidelberg University"
#| message: false
packages(dataRetrieval)
watervilleQ <- read_waterdata_daily(monitoring_location_id = "USGS-04193500",
                          			    parameter_code = c("00060"))

##write.csv(watervilleQ, file="waterville_Q_usgs.csv")
watervilleQ$Rdate <- as.Date(watervilleQ$time, format="%Y-%m-%d")
ttt <- watervilleQ$Rdate >= as.Date("1940-01-01") & 
       watervilleQ$Rdate < as.Date("2024-10-01")

sum(ttt)
dim(wvl.daily)

wvlFlow <- watervilleQ[ttt, c("Rdate", "value")]
names(wvlFlow)[1:2] <- c("Rdate", "Flow")
wvlFlow$Month <- format(wvlFlow$Rdate, format="%Y-%m")  

tt2 <- wvlFlow$Rdate >= as.Date("1982-01-01") 

## comparing data from USGS to data from Heidelberg
##pdf(paste(plotDIR, "Heidelberg_USGS.pdf", sep="/"), height=5, width=5)
par(mar=c(3,3,1,1), mgp=c(1.5,0.5,0), tck=-0.01)
plot(wvlFlow$Flow[tt2] , wvl.daily$FLOW,log="xy", xlab="USGS",
     ylab="Heidelberg")
##dev.off()
```

The USGS data are quite different from the Heidelberg data -- inquiry needed

## Precipitation Data

```{r}
#| label: precip
maumee_precip  <- as_tibble(read.csv(paste(dataDIR, "ERA_Maumee_Daily.csv",
                            					     sep="/"),
                                     header=T))
maumee_precip2  <- as_tibble(read.csv(paste(dataDIR,
                                            "ERA_Maumee_Mon_1940_2022.csv",
                            					     sep="/"),header=T))
## monthly from 1940


names(maumee_precip2)[2:3] <- c("precip_day", "precip_mon")
names(maumee_precip)[2] <- "precip"
maumee_precip$Year <- as.numeric(substring(maumee_precip$Date, 1, 4))
maumee_precip$Month <- as.numeric(substring(maumee_precip$Date, 5, 6))
maumee_precip$Day <- as.numeric(substring(maumee_precip$Date, 7, 8))
maumee_precip$RDate <- as.Date(as.character(maumee_precip$Date),
                      				 format="%Y%m%d")
precip_melton <- melt(maumee_precip, id=c("RDate", "Month","Year", "Month"),
            			     measure.var=c("precip"))
precip_monthly <- dcast(precip_melton, Year+Month ~ variable, mean)

  ## creating time-serie objects
  ## TP
TS_precip_daily <-  ts(maumee_precip$precip, start=c(1979,1, 1), freq=365.25)
TS_precip_monthly <-  ts(precip_monthly$precip, start=c(1979,1), freq=12)
TS_precip_monthly2 <-  ts(maumee_precip2$precip_day, start=c(1940,1), freq=12)

```

## Creating Time-Series

Creating daily and monthly mean time series of variables of interest

```{r}
#| label: time_series
      ## monthly means, try to use median and see if the outcome changes
wvl.daily$TN <- wvl.daily$TKN + wvl.daily$NO23
wvl.daily$SRPvTP <- wvl.daily$SRP / wvl.daily$TP
wvl.daily$TNvTP <- wvl.daily$TN / wvl.daily$TP
wvl.daily$SRPvTP <- wvl.daily$SRP / wvl.daily$TP
wvl.daily$TPLoad <- wvl.daily$TP * wvl.daily$FLOW * 2.44657558 ## to kg/day
wvl.daily$SRPLoad <- wvl.daily$SRP * wvl.daily$FLOW * 2.44657558 ## to kg/day
wvl.daily$TNLoad <- wvl.daily$TN * wvl.daily$FLOW * 2.44657558 ## to kg/day
wvl.daily.melton <- melt(wvl.daily, 
                         id=c("Rdate", "Month","Year", "yrmn"),
                         measure.var=c("FLOW", "TP","TSS","SRP","NO23","TKN",
      					                       "TN", "SRPvTP", "TNvTP", "TPLoad",
                                       "SRPLoad", "TNLoad"))

wvl.monthly <- dcast(wvl.daily.melton, Year+Month+yrmn ~ variable, mean)

      ## creating time-serie objects
      ## TP
TP_ts_daily <-  ts(wvl.daily$TP, start=c(1982,1, 1), freq=365.25)
TP_ts_monthly <-  ts(wvl.monthly$TP, start=c(1982,1), freq=12)

TPLoad_ts_daily <-  ts(wvl.daily$TPLoad, start=c(1982,1, 1), freq=365.25)
TPLoad_ts_monthly <-  ts(wvl.monthly$TPLoad, start=c(1982,1), freq=12)
      
      ## SRP
SRP_ts_daily <- ts(wvl.daily$SRP, start=c(1982,1,1), freq=365.25)
SRP_ts_monthly <- ts(wvl.monthly$SRP, start=c(1982,1), freq=12)

SRPLoad_ts_daily <- ts(wvl.daily$SRPLoad, start=c(1982,1, 1), freq=365.25)
SRPLoad_ts_monthly <- ts(wvl.monthly$SRPLoad, start=c(1982,1), freq=12)

  ## Flow -- in cubic meters per second

Flow_ts_monthly <- ts(wvl.monthly$FLOW*0.028316847, start=c(1982,1),
                      freq=12)
Flow_ts_daily <-  ts(wvl.daily$FLOW*0.028316847, start=c(1982,1,1),
                     freq=365.25)

## Using USGS data from 1940-2024-10
wvlFlow_melton <- melt(wvlFlow, id=c("Rdate","Month"),
                       			    measure.var="Flow")

wvlFlow_m <- dcast(wvlFlow_melton, Month ~ variable, mean)

Flow_ts_monthly_usgs <- ts(wvlFlow_m$Flow*0.028316847, start=c(1940,1),
                          freq=12)
Flow_ts_daily_usgs <-  ts(wvlFlow$Flow*0.028316847, start=c(1940,1,1),
                         freq=365.25)

      ## TN
TN_ts_monthly <- ts(wvl.monthly$TN, start=c(1982,1), freq=12)
TN_ts_daily <-  ts(wvl.daily$TN, start=c(1982,1,1), freq=365.25)

TNLoad_ts_monthly <- ts(wvl.monthly$TNLoad, start=c(1982,1), freq=12)
TNLoad_ts_daily <-  ts(wvl.daily$TNLoad, start=c(1982,1,1), freq=365.25)

  ## NO23
NO23_ts_monthly <- ts(wvl.monthly$NO23, start=c(1982,1), freq=12)
NO23_ts_daily <-  ts(wvl.daily$NO23, start=c(1982,1,1), freq=365.25)
      ## TSS
TSS_ts_monthly <- ts(wvl.monthly$TSS, start=c(1982,1), freq=12)
TSS_ts_daily <-  ts(wvl.daily$TSS, start=c(1982,1,1), freq=365.25)

      ## TNvTP
TNvTP_ts_monthly <- ts(wvl.monthly$TNvTP, start=c(1982,1), freq=12)
TNvTP_ts_daily <-  ts(wvl.daily$TNvTP, start=c(1982,1,1), freq=365.25)

      ## DRPvTP
DRPvTP_ts_monthly <- ts(wvl.monthly$SRPvTP, start=c(1982,1), freq=12)
DRPvTP_ts_daily <-  ts(wvl.daily$SRPvTP, start=c(1982,1,1), freq=365.25)
```

## Saving

Saving clean data for STL analysis in Part II

```{r}
#| label: saving
#| 

save(maumee_precip, maumee_precip2, TS_precip_daily, TS_precip_monthly, TS_precip_monthly2, 
     wvlFlow, wvl.daily, file=paste(dataDIR, "savedWVL.RData", sep="/"))

save(TP_ts_daily, TP_ts_monthly, TPLoad_ts_daily, TPLoad_ts_monthly, SRP_ts_daily,
     SRP_ts_monthly, SRPLoad_ts_daily, SRPLoad_ts_monthly, Flow_ts_monthly, Flow_ts_daily, 
     wvlFlow_melton, wvlFlow_m, Flow_ts_monthly_usgs, Flow_ts_daily_usgs, TN_ts_monthly,
     TN_ts_daily, TNLoad_ts_monthly, TNLoad_ts_daily, NO23_ts_monthly, NO23_ts_daily,
     TSS_ts_monthly, TSS_ts_daily, TNvTP_ts_monthly, TNvTP_ts_daily, DRPvTP_ts_monthly,
     DRPvTP_ts_daily, file=paste(dataDIR, "savedLoadings.RData", sep="/"))

```

## End of Part I
